Predicting the effects of character displacement on food-web dynamics

### 2019-01-23 manuscript_ECD_model.tex
update repo with many changes

\emph{Keywords}: competition; eco-evolutionary dynamics; ecological
speciation; three-spine stickleback; zooplankton

@@

Ecological character displacement is a key driver of diversification and
adaptive radiation; however, there has been little work examining the
consequences of character displacement on the dynamics of ecological
communities. Given the growing recognition for the role of evolution in
structuring ecological communities, understanding how major evolutionary
processes impact ecological dynamics is increasingly important. Here, we
build a mathematical model to predict how exploitative competition
between two consumers for two resources affects the evolution of
consumer attack rates, and, in turn, consumer-resource population
dynamics. Our model suggested that increasing ecological character
displacement decreases the equilibrium density of resources, and with
sufficient divergence, results in more oscillatory (destabilizing)
consumer-resource dynamics. These results suggest that ecological
character displacement can eventually have a negative feedback on
food-web persistence in the absence of other processes.

@@

Ecological character displacement is thought to be a key evolutionary
process generating biodiversity (Dolph Schluter 2000; D. W. Pfennig and
Pfennig 2010; but see Stuart and Losos 2013). Ecological character
displacement describes the ``process of phenotypic evolution in a
species generated or maintained by {[}exploitative{]} resource
competition with one or more coexisting species'' (Dolph Schluter 2000).
Over the past 40 years, a large body of theoretical (e.g. Lawlor and
Smith 1976; P A Abrams 1986; Doebeli 1996; Taper and Chase 1985) and
empirical (reviewed in: Dolph Schluter 2000; Dayan and Simberloff 2005;
Stuart and Losos 2013) work has been generated to understand the
scenarios under which exploitative competition for resources leads to
the divergence of consumer foraging traits. The general result emerging
from these studies is that, if resources are nutritionally substitutable
(Peter A Abrams 1987; Fox and Vasseur 2008) and there is no other strong
source of density dependence acting on consumers (P A Abrams 1986), then
resource competition drives the divergence of consumer foraging traits
{[}Lawlor and Smith (1976); Taper and Chase (1985)). This process is not
simply driven by ecological differences, but creates an eco-evolutionary
feedback that drives further differentiation. This insight was clearly
by theoretical models that highlighted the importance of explicitly
including resource dynamics as a mediator of competition in driving
evolutionary change (Lawlor and Smith 1976; P A Abrams 1986; Taper and
Chase 1985).

Although character displacement is the result of an eco-evolutionary
feedback, the ecological consequences of this process have received
surprisingly little attention (Germain et al. 2018). Perhaps this is
because early theoretical work suggested that character displacement was
a potent stabilizing force in ecological communities (Lawlor and Smith
1976). This ecological outcome is also an intuitive one ---character
divergence reduces interspecific competition between consumers,
promoting coexistence. However, inferences about the effect of character
displacement, like the process itself, can only be made when comparing
communities in the presence and absence of competition (criteria \#3 for
character displacement, D Schluter and McPhail 1992; Dolph Schluter
2000). Current inferences about the stabilizing effects of consumer
divergence are only comparisons about communities where there are two
consumers, and even though, the focus is only on the stable coexistence
of consumer species.

The consequence of this eco-evolutionary feedback for consumer
coexistence has been well characterized (Lawlor and Smith 1976;
Germain's work). Ecological character displacement has typically been
considered to be a stabilizing factor in communities (Lawlor and Smith
1976). This is because divergence in foraging traits is thought to
reduce the strength of interspecific competition between consumers.
Reducing the strength of interspecific competition relative to
intraspecific competition is a stabilizing force in consumer-resource
systems. Consumer specialization on distinct resources suggests that
there are niche differences between consumers. Mathematically, this is
also often assumed to be the case, because using techniques such as
evolutionary invasion analysis describe the ability for a consumer with
a mutant allele to increase in abundance when rare, a demographic
signature of niche differentiation. Therefore, if both consumers are
able to invade when rare, then this suggests that they can coexist.

Character displacement involves evolutionary shifts in character states
in response to competition (D Schluter and McPhail 1992; Dolph Schluter
2000). Theoretically, this is determined by comparing character changes
in the presence and absence of a competitor. Empirically, this is
determined by meeting several criteria that rule out other possible
explanations (D Schluter and McPhail 1992; Dolph Schluter 2000).
Importantly, current inferences about the effect of character
displacement on community stability do not meet this critera. For
example, the inference that character displacement enhances species
coexistence is only made in a scenario where consumers are competing.
This type of stability is important, but it does not enable us to
examine the effect of character displacement on community stability. As
a consequence, I argue that the effect of character displacement on
community stability remains an open question.

Indeed, ecological character displacement often results in the evolution
of characters that enhance a consumer's ability to capture resources,
which tends to destabilize consumer-resource interactions (Murdoch et
al. 2003; McCann 2011). This occurs because increases in a consumer's
attack rate results in the resource being suppressed well below its
carrying capacity, which if sufficient enough, can generate
oscillations, and hence less stable, consumer-resource dynamics
(Rosenzweig 1971; Murdoch et al. 2003). Experimental decreases in
interaction strength between consumers and resources has been shown to
promote coexistence between consumers and resources (Luckinbill 1973).
Moreover, there is broad support that specialized consumers, since they
are tightly coupled to their resources, tend to exhibit
consumer-resource cycles, whereas this is not the case for generalists
(Murdoch et al. 2002). Therefore, although evolutionary responses to
exploitative competition may reduce the strength of interactions between
consumers, they ultimately increase the strength of interactions with
their resources which can destabilize the direct consumer-resource
interaction (rather than the indirect consumer-consumer interaction).
While prior models of character displacement have included more
realistic functional responses and spatial relationships between
consumers and resources, they have focused more on whether these factors
alter patterns of character displacement rather than the effects of
character displacement on the stability of the ecological community. As
a consequence, it is currently unclear whether ecological character
displacement has a net stabilizing or destabilizing effect on ecological
communities.

Here, I address this knowledge gap by building a mathematical model to
examine how ecological character displacement affects consumer-resource
dynamics in a food web context. I address two question: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? I show that the
consequences of character displacement on food-web dynamics are not
intuitive and depend critically on the foraging scenario and tradeoff in
consumer attack rates.

@@

\subsection{Model of resource
competition}\label{model-of-resource-competition}

@@

\[\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2\]
\[\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2\]
\[\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1\]
\[\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2\]

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ji}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ji}(R_i)\) represents consumer \(j\)'s feeding rate
(i.e.~functional response) on resource \(i\). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates) rather than if
resources are better characterized by a continuous trait distribution
(e.g.~seed size, see Taper and Chase (1985) for an example). In the
scenario where consumer feeding rates increase linearly with resource
density, such that \(F_{ji}(R_i)=a_{ji}R_i\) where \(a_{ji}\) is the
attack rate of consumer \(j\) on resource \(i\), then this model becomes
MacArthur's model of resource competition (MacArthur (1972)). The
evolution of consumer attack rates in this model (and several
extensions) have been analyzed in detail (Lawlor and Smith (1976); P A
Abrams (1986)). The general result of these analyses is that divergent
character displacement in attack rates is an inevitable outcome of
resource competition.

Here, I examine the effects of this inevitable divergence in attack
rates on food-web dynamics. In particular, I examine how divergence in
attack rates affect resource densities as well as the local stability of
both resources and consumers. I do this by (i) deriving analytical
expressions for the relationship between attack rates and food-web
dynamics in different foraging scenarios and (ii) simulating the effects
of competition on the eco-evolutionary dynamics of consumers and
resources. For these simulations, I used an Adaptive Dynamics approach.
At each evolutionary time step, I created a mutant consumer by randomly
choosing a consumer and modifying its attack rate on one resource by
either subtracting or adding a small constant with equal probability.
The mutant's attack rate on the other resource was determined by the
shape of the tradeoff (details below). Assuming this mutant consumer is
rare, I then determined whether the mutant had higher relative fitness
than the resident consumer, and thus could invade and replace the
resident consumer. If the mutant was able to invade, I updated the
attack rate of the resident consumer to the mutant attack rate and
allowed consumer and resource dynamics to reach a steady state. I then
repeated the simulation up to 10,000 times, which was sufficient for
consumers to either reach an evolutionary stable strategy (ESS, Smith
and Price (1973)) or an evolutionary limit (e.g.
\(\frac{a_{ii}}{a_{ii}+a_{ij}}\) is constrained to a maximum of 1 and
minimum of 0).

\subsection{Resource densities}\label{resource-densities}

I start by examing the effects of character displacement on the density
of resources at equilibrium, which will set the stage for a later
analysis of food-web stability. In particular, I examine two different
foraging scenarios: one where consumers forage for both resources
simultaneously (e.g. MacArthur (1972)) and one where they cannot because
resources occur in distinct habitats (e.g. Lawlor and Smith (1976); K S
McCann, Rasmussen, and Umbanhowar (2005)). I say consumers have
undergone divergent character displacement if there evolved attack rates
are more specialized, defined as \(|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|\),
when evolving with a competing consumer.

\textbf{Consumers forage for both resources simultaneously}: An inherent
assumption in MacArthur's model of resource competition is that
consumers can forage for both resources simultaneously. In essence, this
implies that both resources occur within the same habitat. To gain
analytical insight to this scenario, I assume that resources are
equivalent (\(r=r_1=r_2\) and \(K=K_1=K_2\)) as well as consumers
(\(e=e_{11}=e_{12}=e_{21}=e_{22}\); \(m=m_1=m_2\)), except that consumer
attack rates are mirror images of each other (\(a_{ii}=a_{11}=a_{22}\);
\(a_{ij}=a_{12}=a_{21}\)). While this scenario is arguably simplistic,
it isolates the effect of character displacement rather than any
dependence on inherent differences in consumers or resources. I show
that when both consumers and resources are present that the density of
resources at equilibrium are equivalent (\(R=R_1=R_2\)) and are
determined by the following equation (\emph{Mathematica} file with
derivation available upon request):

\[\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}\] Note that a key
determinant of resource density is the consumer's total attack rate,
\(a_{ii}+a_{ij}\). Since the total attack rate is in the denominator,
this implies that resource densities will decrease if evolution favors
an increase in total attack rate. I assume that consumers are not
`Darwinian demons' (Law (1979)) and thus the evolution of their attack
rates are subject to certain constraints. I modelled these constraints
such that \(\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1\), where
\(A\) is the total investment in attack rates and \(n\) describes the
shape of the tradeoff (Sargent and Otto (2006)). This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{ii}\) and \(a_{ij}\) are higher, on average, than
the extremes (when \(n>1\), green line in Fig. \ref{fig:Tradeoff}) or,
conversely, where the two extremes are higher, on average, than
intermediate investments (when \(n<1\), orange line Fig.
\ref{fig:Tradeoff}). When \(n=1\), the tradeoff function is linear, and
all combinations of \(a_{ii}\) and \(a_{ij}\) have the same total attack
rate (blue line in Fig. \ref{fig:Tradeoff}).

\begin{figure}
\centering
\includegraphics{ECD_Supp_Mat_files/figure-latex/Tradeoff-1.pdf}
\caption{\label{fig:Tradeoff}Tradeoff forms for consumer attack rates.}
\end{figure}

Previous work has shown that consumers undergo divergent character
displacement regardless of the shape of the tradeoff (Lawlor and Smith
(1976); P A Abrams (1986)). Interestingly, I find that the shape of the
tradeoff qualitatively affects the relationship between character
displacement and resource density (Fig. \ref{fig:MacArthur_Resources}).
For example, if consumer's are constrained by a linear tradeoff (blue
lines), then there is no net change in total attack rate (Fig.
\ref{fig:MacArthur_Resources}A) and character displacement has no effect
on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the
tradeoff is concave down (green lines), then resource abundances can
actually increase under character displacement (Fig.
\ref{fig:MacArthur_Resources}B). This is because the total attack rate
of consumers is maximized at intermediate values (\(a_{ii}=a_{ij}\)) and
decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A).
Note that detecting this effect in nature would likely be difficult,
since a concave-down tradeoff results in relatively small displacement
relative to other tradeoff shapes (Lawlor and Smith (1976); green points
in Fig. \ref{fig:MacArthur_Resources}A,B). When the tradeoff is concave
up (orange lines), character displacement suppresses resource densities
due to the increase in total attack rates (Fig.
\ref{fig:MacArthur_Resources}A,B). Although the equation I derived for
resource density was for the scenario where both consumers and both
resources are present, it predicts well the density of resources when a
single consumer reaches its ESS (triangles on respective colored lines
in Fig. \ref{fig:MacArthur_Resources}B). This is because a single
consumer evolves to be a generalist that has equal attack rates on each
resource (triangles at 0.5 along x-axis in Fig.
\ref{fig:MacArthur_Resources}A), resulting in equivalent resource
densities.

\begin{figure}
\centering
\includegraphics{ECD_Supp_Mat_files/figure-latex/MacArthur_Resources-1.pdf}
\caption{\label{fig:MacArthur_Resources}Effect of character displacement
on total attack rates (A) and resource densities (B) when consumers can
forage for both resources simultaneously.}
\end{figure}

\textbf{Consumers \emph{cannot} forage for both resources
simultaneously}: Most models of character displacement have assumed a
scenario where consumers can forage for both resources simultaneously
(e.g. Taper and Chase (1985); P A Abrams (1986)). This assumption may be
valid for some foraging scenarios (e.g.~Darwin's finches foraging for
seeds); however, in many situations consumers forage for resources that
occur in different habitats, and thus cannot forage for both
simultaneosly. The only character displacement model that I am aware of
that modeled resources in different habitats was one examined by Lawlor
and Smith (1976). This model takes the same form as the previous model,
except now the consumer's feeding rate takes the form:

@@

where \(w_{ji}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(j\) is found (i.e.~habitat
preference). Note that since \(w_{ji}\) is a proportion that
\(w_{jj}=1-w_{ji}\). As with attack rates, I assume that consumer
habitat preferences are mirror images of each other
(\(w=w_{11}=w_{22}\)).

Lawlor and Smith (1976) found that consumers still underwent divergent
character displacement regardless of whether consumers could or could
not forage for resources simultaneously. I find that the foraging
scenario qualitatively affects the relationship between character
displacement and resource densities. To demonstrate this, note that
again resource densities are equivalent at equilibrium (\(R=R_1=R_2\)
when both consumers and resources present), but are now determined by
the following equation (\emph{Mathematica} file with derivation
available upon request):

@@

This equation implies that if consumers evolve to become specialists on
resources that occur in their preferred habitat (e.g. \(w>0.5\) and
\(a_{ii}>a_{ij}\)), then character displacement will always increase the
effective attack rate of consumers (\(wa_{ii}+(1-w)a_{ij}\)), regardless
of the tradeoff (Fig. \ref{fig:LS_Resources}A). Thus, character
displacement always results in resource suppression if consumers are
competing for resources that occur in different habitats (Fig.
\ref{fig:LS_Resources}B). Note that the shape of the tradeoff can modify
the effect of character displacement. This is not so much due to the
tradeoff affecting the magnitude of displacement (it does, but the
effect is minor), but because the form of the tradeoff affects resource
abundances when a single consumer has reached its ESS (triangles in Fig.
\ref{fig:LS_Resources}B). In contrast, resource densities reach a
similar value when consumers evolve in the presence of a competitor
(circles in Fig. \ref{fig:LS_Resources}B), because character
displacement tends to reach a constraint of complete specialization.

It is worth noting that in this foraging scenario, resource densities
are consistently higher at the single consumer ESS compared to the
predictions I derived for when both consumers are present (deviation of
triangles from respective colored lines in Fig.
\ref{fig:LS_Resources}B). This is likely because consumers actually
evolve to be slightly specialized on the resources that occur in their
non-preferred habitat (deviation of triangles from 0.5 along x-axis in
Fig. \ref{fig:LS_Resources}B).

\begin{figure}
\centering
\includegraphics{ECD_Supp_Mat_files/figure-latex/LS_Resources-1.pdf}
\caption{\label{fig:LS_Resources}Effect of character displacement on
effective attack rates (A) and resource densities (B) when consumers
cannot forage for resources simultaneously.}
\end{figure}

The effect of the foraging scenario on the relationship between
character displacement and resource densities appears to be quite
general. For example, even if I consider a more realistic consumer
functional response (derived by K S McCann, Rasmussen, and Umbanhowar
(2005)):

@@

\subsection{Food-web stability}\label{food-web-stability}

Of the prior models of character displacement, only Lawlor and Smith
(1976) examine the effect of character displacement on the ecological
stability of the system. They conducted a detailed theoretical analysis
of three different metrics of stability: (i) stable coexistence of
consumers, defined as the mutual invasibility of each consumer; (ii)
global stability of consumers, defined as the continued coexistence of
both consumers when resource dynamics are perturbed; and (iii) local
stability, defined as the rate at which the four-species food web
returned to equilibrium after a small perturbation to species
abundances. Note that the first two metrics correspond to the stable
coexistence of consumers, while the third metric measures the stability
of the entire food web. Lawlor and Smith (1976) found that character
displacement enhanced the stable coexistence of consumers, yet had a
small negative effect on the local stability of the food web.

Lawlor and Smith (1976)'s work gave fundamental insight to the effect of
character displacement on ecological stability; however, none of their
stability analyses compared food webs with and without a competing
consumer. Such comparisons are necessary for inferring the effects of
ecological character displacement, a point that has been made clear in
the criteria to demonstrate character displacement (Dolph Schluter
(2000); D Schluter and McPhail (1992)). In fact, the first two metrics
of stability can only be assessed when both consumers are present
because it examines the effect of divergence on coexistence, which
requires both consumers to be present. In contrast, local stability can
be compared in food webs with and without a competing consumer, and thus
provides a useful benchmark for examining the effects of character
displacement on food-web stability.

Below, I plot the results from the same simulations I did previously
except now local stability
(\(-1\times\operatorname{Re}(\lambda_{max})\)) is on the y-axis (Fig.
\ref{fig:Stability}). In general, I find that resource supression goes
hand-in-hand with local stability, a tendency that has been noted by
others (Murdoch, Briggs, and Nisbet (2003); Kevin S McCann (2011)).
Whenever character displacement results in resource suppression, I
observe a decrease in local stability (Fig. \ref{fig:Stability}). This
is not simply a consequence of having an additional consumer in the
system, but emerges from the eco-evolutionary feedback between character
displacement and resource suppression (compare initial (small points)
with evolved strategies (large points), Fig. \ref{fig:Stability}). Note
that I reach the same fundamental conclusions when I incorporated a more
realistic functional response for the consumers (Fig.
\ref{fig:Stability_McCann} in Appendix).

>>>>

\emph{Keywords}: competition; eco-evolutionary dynamics;
consumer-resource interactions; adaptive radiation; ecological community

@@

@@

Ecological character displacement is thought to be a key evolutionary
process generating biodiversity (Dolph Schluter 2000; D. W. Pfennig and
Pfennig 2010; but see Stuart and Losos 2013). Ecological character
displacement describes the ``process of phenotypic evolution in a
species generated or maintained by {[}exploitative{]} resource
competition with one or more coexisting species'' (Dolph Schluter 2000).
Over the past 40 years, a large body of theoretical (e.g. Lawlor and
Smith 1976; P A Abrams 1986; Doebeli 1996; Taper and Chase 1985) and
empirical (reviewed in: Dolph Schluter 2000; Dayan and Simberloff 2005;
Stuart and Losos 2013) work has been generated to understand the
scenarios under which exploitative competition for resources leads to
the divergence of consumer foraging traits. The general result emerging
from these studies is that, if resources are nutritionally substitutable
(Peter A Abrams 1987; Fox and Vasseur 2008) and there is no other strong
source of density dependence acting on consumers (P A Abrams 1986), then
resource competition drives the divergence of consumer foraging traits
{[}Lawlor and Smith (1976); Taper and Chase (1985)). This process is not
simply driven by ecological differences, but creates an eco-evolutionary
feedback that drives further differentiation. This key insight was made
by theoretical models that explicitly included resource dynamics as a
mediator of competition in driving evolutionary change (Lawlor and Smith
1976; P A Abrams 1986; Taper and Chase 1985).

Although explicit models of resources gave key insights to the evolution
of character displacement, the ecological feedback onto
consumer-resource dynamics have received surprisingly little attention.
This is likely because character displacement has been primarily studied
through the lens of competition and coexistence theory {[}Lawlor1976;
Germain2018{]}. For example, early theoretical work by Lawlor and Smith
(1976) showed that ecological character displacement promotes
coexistence by favoring specialized consumers that experience reduced
interspecific competition. On the other hand, specialization often
increases a consumer's ability to capture a specific resource. A well
known consequence of increased foraging efficiency is resource
suppression, which if sufficient enough, can generate oscillations, and
hence less stable, consumer-resource dynamics {[}Rosenzweig 1971;
Murdoch et al. 2003; McCann 2011{]}. NEED A LAST SENTENCE TO WRAP IT UP,
HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL
INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY.
PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A
FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY
BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT
SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP
PREDATOR (I.E. NO CONSUMERS\ldots{}LOOK AT OLD TEXT FOR WORDING
SUGGESTIONS.)

Here, I address this knowledge gap by building a mathematical model to
examine how ecological character displacement affects consumer-resource
dynamics in a food-web context. I address two questions: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? To test the
generality of these ecological effects, I examined three different
foraging scenarios: (i) resources occur in the same habitat and
consumers exhibit a linear functional response (e.g. MacArthur (1972);
Lawlor and Smith (1976)); (ii) resources occur in different habitats and
consumers exhibit a linear functional response (e.g. Lawlor and Smith
(1976)); (iii) resources occur in distinct habitats and consumers
exhibit a more realistic functional response (K S McCann, Rasmussen, and
Umbanhowar 2005). I also explore the contingency of these effects to
different assumptions about evolutionary trade-offs in consumer foraging
traits. I focus on the ecological side of these dynamics because the
evolutionary aspects have been well characterized (Lawlor and Smith
1976; Taper and Chase 1985; P A Abrams 1986). I show that the
consequences of character displacement on food-web dynamics are not
always intuitive and depend on the foraging scenario and tradeoff in
consumer traits.

@@

\subsection{Underlying Food-web
Dynamics}\label{underlying-food-web-dynamics}

@@

\[\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2,\]
\[\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2,\]
\[\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1,\]
\[\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2,\]

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ji}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ji}(R_i)\) represents consumer \(j\)'s feeding rate
(i.e.~functional response) on resource \(i\). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates) rather than if
resources are better characterized by a continuous trait distribution
(e.g.~seed size, see Taper and Chase (1985) for an example). In the
scenario where consumer feeding rates increase linearly with resource
density, such that \(F_{ji}(R_i)=a_{ji}R_i\) where \(a_{ji}\) is the
attack rate of consumer \(j\) on resource \(i\), then this model becomes
MacArthur's model of resource competition (MacArthur (1972)). The
evolution of consumer attack rates in this model (and several
extensions) have been analyzed in detail (Lawlor and Smith (1976); P A
Abrams (1986)), with the general result being divergent character
displacement. I say consumers have undergone divergent character
displacement if there evolved attack rates are more specialized, defined
as \(|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|\), when evolving with a
competing consumer. I focus here on the ecological consequences of this
adaptive divergence.

To investigate the ecological consequences, I studied the effect of
adaptive divergence in consumers on resource densities and the stability
of the entire food web at equilibrium. An equilibrium is reached when
the four rates of change in the above equations are 0, and solving the
system at this point gives equilibrium densities for each resource
(\(\hat R_1,\hat R_2\)) and consumer (\(\hat C_1,\hat C_2\)). At the
coexistence equilibrium (i.e., where
\(\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0\)), I calculated resource
densities and performed a linear stability analysis (Otto and Day 2007).
This stability analysis derives the dominant (largest in absolute value)
eigenvalue, \(\lambda\), which determines whether (and how readily) the
food web, when perturbed a small amount, will return to equilibrium (see
supplementary material for derivation). Together, these two measures
indicate how ecological character displacement affects resource
densities and food-web stability.

I examine the effects of this inevitable divergence in attack rates on
food-web dynamics by (i) deriving analytical expressions for the
relationship between attack rates and food-web dynamics and (ii)
simulating the effects of competition on the eco-evolutionary dynamics
of consumers and resources. To gain analytical insight, I assume that
resources are equivalent (\(r=r_1=r_2\) and \(K=K_1=K_2\)) as well as
consumers (\(e=e_{11}=e_{12}=e_{21}=e_{22}\); \(m=m_1=m_2\)), except
that consumer attack rates are mirror images of each other
(\(a_{ii}=a_{11}=a_{22}\); \(a_{ij}=a_{12}=a_{21}\)). While this
scenario is arguably simplistic, it allows me to isolate the effects of
ecological character displacement.

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics
approach. At each evolutionary time step, I created a mutant consumer by
randomly choosing a consumer and modifying its attack rate on one
resource by either subtracting or adding a small constant (0.001 in the
following simulations) with equal probability. The mutant's attack rate
on the other resource was determined by a tradeoff, such that
\(\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1\), where \(A\) is the
total investment in attack rates and \(n\) describes the shape of the
tradeoff (Sargent and Otto 2006). This function has the useful property
that it differentiates between cases where intermediate combinations of
\(a_{ii}\) and \(a_{ij}\) are higher, on average, than the extremes
(when \(n>1\), green line in Fig. \ref{fig:Tradeoff}) or, conversely,
where the two extremes are higher, on average, than intermediate
investments (when \(n<1\), orange line Fig. \ref{fig:Tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{ii}\) and \(a_{ij}\) have the same total attack rate (blue line in
Fig. \ref{fig:Tradeoff}). Assuming the mutant consumer was rare, I then
determined whether the mutant had higher relative fitness than the
resident consumer, and thus could invade and replace the resident
consumer. If the mutant was able to invade, I updated the attack rate of
the resident consumer to the mutant attack rate and allowed consumer and
resource dynamics to reach a steady state. I then repeated the
simulation up to 10,000 times, which was sufficient for consumers to
either reach an evolutionary stable strategy (ESS, Smith and Price 1973)
or an evolutionary limit (e.g. \(\frac{a_{ii}}{a_{ii}+a_{ij}}\) is
constrained to a maximum of 1 and minimum of 0).

\begin{figure}
\centering
\includegraphics{manuscript_ECD_model_files/figure-latex/Tradeoff-1.pdf}
\caption{\label{fig:Tradeoff}Tradeoff forms for consumer attack rates.}
\end{figure}

\subsection{Consumers Forage For Both Resources
Simultaneously}\label{consumers-forage-for-both-resources-simultaneously}

When both consumers and resources are present, the density of resources
at equilibrium are equivalent (\(\hat R = \hat R_1 = \hat R_2\)) and are
determined by the following equation (derivation in supplementary
material):

\[\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}\] A key determinant of
resource density in this model is the consumer's total attack rate,
\(a_{ii}+a_{ij}\). Therefore, the ecological consequences of character
displacement depend on how the tradeoff influences the evolution of
consumer attack rates.

My simulations show that the shape of the attack-rate tradeoff
qualitatively affects the relationship between character displacement
and resource density (Fig. \ref{fig:MacArthur_Resources}). For example,
if consumer's are constrained by a linear tradeoff (blue lines), then
there is no net change in total attack rate (Fig.
\ref{fig:MacArthur_Resources}A) and character displacement has no effect
on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the
tradeoff is concave down (green lines), then resource abundances can
actually increase under character displacement (Fig.
\ref{fig:MacArthur_Resources}B). This is because the total attack rate
of consumers is maximized at intermediate values (\(a_{ii}=a_{ij}\)) and
decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A).
When the tradeoff is concave up (orange lines), character displacement
suppresses resource densities due to the increase in total attack rates
(Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived
for resource density was for the scenario where both consumers and both
resources are present, it accurately predicts the density of resources
when a single consumer reaches its ESS (triangles on respective colored
lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single
consumer evolves to be a generalist that has equal attack rates on each
resource (triangles at 0.5 along x-axis in Fig.
\ref{fig:MacArthur_Resources}A), resulting in equivalent resource
densities.

The effect of character displacement on resource density closely matches
its impact on food-web stability (Fig. \ref{fig:Stability}). For
example, when character displacement results in resource suppression
(orange), there is a corresponding decrease in stability. Similarly, if
character displacement does not influence resource densities (blue),
then there is no corresponding effect on stability. Curiously, this
match between impact on resources and stability breakdowns with the
concave down tradeoff (green).

\begin{figure}
\centering
\includegraphics{manuscript_ECD_model_files/figure-latex/MacArthur_Resources-1.pdf}
\caption{\label{fig:MacArthur_Resources}Effect of character displacement
on total attack rates (A) and resource densities (B) when consumers can
forage for both resources simultaneously.}
\end{figure}

\includegraphics{Figure_ECD_MacArthur.pdf}

\subsection{\texorpdfstring{Consumers \emph{cannot} forage for both
resources
simulanteously}{Consumers cannot forage for both resources simulanteously}}\label{consumers-cannot-forage-for-both-resources-simulanteously}

The only character displacement model that I am aware of that modeled
resources in different habitats was one examined by Lawlor and Smith
(1976). This model takes the same form as the previous model, except now
the consumer's feeding rate takes the form:

@@

where \(w_{ji}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(j\) is found (i.e.~habitat
preference). Note that since \(w_{ji}\) is a proportion that
\(w_{jj}=1-w_{ji}\). As with attack rates, I assume that consumer
habitat preferences are mirror images of each other
(\(w=w_{11}=w_{22}\)). In Lawlor and Smith (1976)'s detailed
evolutionary analysis of this model, they again always observe divergent
character displacement.

To examine the ecological consequences of this model, I again derive the
equilibrium solution when both consumers and both resources are present.
Again, resource densities are equivalent at this equilibrium
(\(\hat R = \hat R_1 = \hat R_2\) when both consumers and resources
present), but are now determined by the following equation (derivation
in supplementary material):

@@

This equation implies that if consumers evolve to become specialists on
resources that occur in their preferred habitat (e.g. \(w>0.5\) and
\(a_{ii}>a_{ij}\)), then the effective attack rate of consumers
(\(wa_{ii}+(1-w)a_{ij}\)) will always increase, regardless of the
tradeoff (Fig. \ref{fig:LS_Resources}A). Thus, character displacement
always results in resource suppression if consumers are competing for
resources that occur in different habitats (Fig.
\ref{fig:LS_Resources}B). Note that the shape of the tradeoff can modify
the effect of character displacement. This is not so much due to the
tradeoff affecting the magnitude of displacement (it does, but the
effect is minor), but because the form of the tradeoff affects resource
abundances when a single consumer has reached its ESS (triangles in Fig.
\ref{fig:LS_Resources}B). In contrast, resource densities reach a
similar value when consumers evolve in the presence of a competitor
(circles in Fig. \ref{fig:LS_Resources}B), because character
displacement tends to reach a constraint of complete specialization. It
is worth noting that in this foraging scenario, resource densities are
consistently higher at the single consumer ESS compared to the
predictions I derived for when both consumers are present (deviation of
triangles from respective colored lines in Fig.
\ref{fig:LS_Resources}B). This is likely because consumers actually
evolve to be slightly specialized on the resources that occur in their
non-preferred habitat (deviation of triangles from 0.5 along x-axis in
Fig. \ref{fig:LS_Resources}B).

The effect of character displacement on resource densities qualitatively
matches its negative effect of food-web stability (Fig.
\ref{fig:Stability}). This is not simply a consequence of having an
additional consumer in the system, but emerges from the eco-evolutionary
feedback between character displacement and resource suppression
(compare initial (small points) with evolved strategies (large points),
Fig. \ref{fig:Stability}). For example, when the tradeoff is concave up
(orange), then the four-species (small circle) food web actually starts
off as less stable than the three-species one (small triangle), but the
eco-evolutionary dynamics cause a switch in the pattern once they reach
their evolutionary stable strategies (large points).

\begin{figure}
\centering
\includegraphics{manuscript_ECD_model_files/figure-latex/LS_Resources-1.pdf}
\caption{\label{fig:LS_Resources}Effect of character displacement on
effective attack rates (A) and resource densities (B) when consumers
cannot forage for resources simultaneously.}
\end{figure}

\includegraphics{Figure_ECD_LS.pdf}

\subsection{Adding A More Realistic Functional
Response}\label{adding-a-more-realistic-functional-response}

Thus far I have consider a simple type I functional response of
consumers foraging for resources in either the same or distinct
habitats.\\
For example, even if I consider a more realistic consumer functional
response (derived by K S McCann, Rasmussen, and Umbanhowar (2005)):

@@
